Further Evidence for Modified Spin-down in Sun-like Stars:\\Pile-ups in the Temperature--Period Distribution

### 2021-12-17 src/ms.tex
jvs comments

For example, efforts to calibrate gyrochronology relations using Kepler asteroseismic targets revealed tension with relations calibrated to open clusters and found that rotation periods could not be described by a single power-law relation with age \citep{Angus2015}. This tension, at least in part, is due to the fact that standard gyrochronology models are unable to account for the anomalously rapid rotation rates of stars older than the Sun, leading to the suggestion that stars with masses $\gtrsim$~1\msun experience a phase of weakened magnetic braking \citep[WMB,][]{vanSaders2016}. Forward modeling simulations of the observed Kepler rotation period distribution also provided support for the WMB hypothesis over standard spin-down models, in that WMB models are better able to match the observed long-period edge \citep{vanSaders2019}. Those authors also predicted a pile-up of stars along the long-period edge, which they hypothesized could not be seen in the \citet{McQuillan2014} sample due to large errors on \teff in the Kepler Input Catalog \citep[KIC,][]{Brown2011}. While \citet{vanSaders2019} favored the WMB hypothesis to explain observations, those authors were also careful to point out that a long-period edge can be caused by detection biases, as stars with longer rotation periods (and larger Rossby numbers) have smaller amplitude variations which pose more difficulty to period-detection algorithms.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{./figures/surveys.pdf}
    \caption{The \teff-\prot plane using rotation periods from \citet{McQuillan2014} or, in the case of the CKS sample, \citet{McQuillan2013}, with \teff originating from the source denoted at top. The \citet{McQuillan2014} \teff values originate from the Kepler Input Catalog \citep[KIC,][]{Brown2011} or \citet{Dressing2013} for low-mass stars. The orange point in each panel indicates the Sun's equatorial rotation period, with the errorbar capturing the range of periods measured from the differentially rotating surface.}
    \label{fig:surveys}
\end{figure*}

@@

The short- and long-period pile-ups we examine here naturally produce a dearth of rotators at intermediate periods. This gap is the same period gap noticed by the authors mentioned above, as made apparent when comparing the \lamostmcq sample to the original \citet{McQuillan2014} sample. Moreover, as seen in Figure~\ref{fig:gap}, we recover the gradient in photometric variability across the gap pointed out by \citet{Reinhold2019}. The photometric variability, as measured through the \rper metric published by \citet{McQuillan2014}, reaches a local minimum near the location of the gap. This supports the notion that the cause of the gap is due to changes in the stars themselves, rather than being the result of mixed stellar populations. However, the variability levels on both sides of the gap are not close to the detection limit, as evidenced by the fact that periods are securely detected for stars with similar properties at much lower \rper values. This would suggest that the period gap is not due solely to a detection issue, unless variability levels were to drop precipitously as stars approached the gap. The data presented here provide evidence for the gap being caused by the same phenomena as the short-period pile-up, namely, core-envelope coupling. The decline in photometric variability that coincides with the gap is likely not the cause of it, but may result from a period of altered or decreased magnetic activity that is connected with core-envelope coupling.

@@

    \item The number density of stars on the long-period pile-up declines with \teff, though it remains unclear whether this is due to astrophysics, the Kepler selection function, observational biases, or some combination of effects. The relative fraction of stars on the long-period pile up declines by a factor of $\sim$2 between $\sim$6200~K and $\sim$5800~K.

>>>>

For example, efforts to calibrate gyrochronology relations using Kepler asteroseismic targets revealed tension with relations calibrated to open clusters and found that rotation periods could not be described by a single power-law relation with age \citep{Angus2015}. This tension, at least in part, is due to the fact that standard gyrochronology models are unable to account for the anomalously rapid rotation rates of stars older than the Sun, leading to the suggestion that stars with Ro~$>$~2 experience a phase of weakened magnetic braking \citep[WMB,][]{vanSaders2016}.

Forward modeling simulations of the observed Kepler rotation period distribution also provided support for the WMB hypothesis over standard spin-down models, in that WMB models are better able to match the observed long-period edge \citep{vanSaders2019}. Those authors also predicted a pile-up of stars along the long-period edge, which they hypothesized could not be seen in the \citet{McQuillan2014} sample due to large errors on \teff in the Kepler Input Catalog \citep[KIC,][]{Brown2011}. While \citet{vanSaders2019} favored the WMB hypothesis to explain observations, those authors were also careful to point out that a long-period edge can be caused by detection biases, as stars with longer rotation periods (and larger Rossby numbers) have smaller amplitude variations which pose more difficulty to period-detection algorithms.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{./figures/surveys.pdf}
    \caption{The \teff-\prot plane using rotation periods from \citet{McQuillan2014} or, in the case of the CKS sample, \citet{McQuillan2013} which applied an identical analysis to Kepler Objects of Interest (KOIs), with \teff originating from the source denoted at top. The \citet{McQuillan2014} \teff values originate from the Kepler Input Catalog \citep[KIC,][]{Brown2011} or \citet{Dressing2013} for low-mass stars. The orange point in each panel indicates the Sun's equatorial rotation period, with the errorbar capturing the range of periods measured from the differentially rotating surface. Many of the stars above the long-period pile-up are subgiants which have experienced spin-down due to expansion off the main-sequence, as pointed out in \citet{vanSaders2019}.}
    \label{fig:surveys}
\end{figure*}

@@

The short- and long-period pile-ups we examine here naturally produce a dearth of rotators at intermediate periods. This gap is the same period gap noticed by the authors mentioned above, as made apparent when comparing the \lamostmcq sample to the original \citet{McQuillan2014} sample. Moreover, as seen in Figure~\ref{fig:gap}, we recover the gradient in photometric variability across the gap pointed out by \citet{Reinhold2019}. The photometric variability, as measured through the \rper metric published by \citet{McQuillan2014}, reaches a local minimum near the location of the gap. This supports the notion that the cause of the gap is due to changes in the stars themselves, rather than being the result of mixed stellar populations. However, the variability levels on both sides of the gap are not close to the detection limit, as evidenced by the fact that periods are securely detected for stars with similar properties at much lower \rper values. This would suggest that the period gap is not due solely to a detection issue, unless variability levels were to drop precipitously as stars approached the gap.

The data presented here provide evidence for the gap being caused by the same phenomena as the short-period pile-up, namely, core-envelope coupling. The decline in photometric variability that coincides with the gap is likely not the cause of it, but may result from a period of altered or decreased magnetic activity that is connected with core-envelope coupling. Connections between the short-period pile-up, the decline in photometric variability, and the existence of the intermediate period gap may also be consistent with the suggestion discussed in \citet{Gordon2021} that the gap could be due to a period of accelerated spin-down immediately proceeding the stalling due to core-envelope coupling. In this picture, spin-down slows as angular momentum is transferred from the core to the envelope, which is accompanied by a decline in photometric variability. After some time, spin-down resumes, but at an accelerated rate such that stars cross the gap quickly and are rarely observed there. \textbf{An increase in photometric variability accompanies this period of...}

@@

    \item The number density of stars on the long-period pile-up declines with \teff, in line with predictions from the WMB hypothesis, though it remains unclear whether this observation is due to astrophysics, the Kepler selection function, observational biases, or some combination of effects. The relative fraction of stars on the long-period pile up declines by a factor of $\sim$2 between $\sim$6200~K and $\sim$5800~K.
